Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-02-09 bbbq_article.tex
Use stored results

To determine the percentages of MHC-I and MHC-II epitopes overlap
with TMHs, we used mostly the same analysis as described in \cite{bianchi2017}.
To summarize: from a proteome all possible 9-mers are derived. For each
of these peptides, it is measured if it has been part of a 
TMH (at least overlapping 1 residue), 
as well as if the peptide is a binder to an MHC-I haplotype.
For MHC-II, 15-mers are used, as these are the most occurring
epitope length \cite{bergseng2015different}.

@@

To detect the evolutionary conservation of TMHs, 
we use the following workflow (also shown
in figure \ref{fig:snp_workflow}) to be able to count the
number of SNPs that occur in TMHs, as well as the number of SNPs
that occur in TMHs by chance.

To do so, we use the 
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases:
'gene' to find the gene names of membrane proteins, 
'dbSNP' (\cite{sherry2001dbsnp}) for SNPs associated with those genes
and 'protein' to obtain the sequence of proteins that SNPs act upon.
The 'dbSNP' contains 2 billion (human-only) SNPs, although
it is estimated that approximately ten percent (ranging from 2 to 36 percent) 
are artifacts (\cite{carlson2003additional, cutler2001high, gabriel2002structure, mitchell2004discrepancies, musumeci2010single, reich2003quality}).

@@

% \paragraph{Introduction}
% 
% To detect the evolutionary conservation of TMHs,
% we first selected all unique protein sequences per
% gene from all 45k SARS-CoV-2 proteomes,
% resulting in around 500 unique protein sequences per gene.
% From the unique protein sequences, we made an alignment per gene,
% using the msa (\cite{bodenhofer2015msa}) R package, 
% set up to use ClustalOmega (\cite{thompson1994clustal})
% with the BLOSUM80 amino acid transition matrix (\cite{henikoff1992amino}).
% Per alignment, we used msa to determine the consensus sequence
% and the conservation scores.
% Per consensus sequences, the topology was determined using pureseqtmr.
% 
% \paragraph{Statistics}
% 
% We compare the conservation values for TMH versus non-TMH locations,
% for all membrane proteins combined,
% using a Kolmogorov-Smirnov test, for a significance level of 5 percent.

@@

\begin{figure}[!htbp]
  \includegraphics[height=0.9\textheight]{../bbbq_1_smart/fig_f_tmh_mhc1_2_grid.png}
  \caption{
    Percentage of MHC-I epitopes overlapping with TMHs
    for a human, viral and bacterial proteome.
    The horizontal lines indicate the percentage as expected by chance.
    \richel{Refer to a table with the numbers as expected by chance)}
    See table \ref{tab:tmh_binders_mhc1} for the TMH and epitope counts.
    \richel{Epitope prediction by EpitopePrediction.}.
    Note that for smaller proteomes a percentage of zero is likelier.
    \richel{use of top 2\% binders}
  }
  \label{fig:1}
\end{figure}

@@

\begin{figure}[!htbp]
  \includegraphics[height=0.9\textheight]{../bbbq_1_smart/fig_f_tmh_mhc2_2_grid.png}
  \caption{
    Percentage of MHC-II epitopes overlapping with TMHs
    for a human, viral and bacterial proteome.
    The horizontal lines indicate the percentage as expected by chance.
    \richel{Refer to a table with the numbers as expected by chance)}
    \richel{Epitope prediction by NetMHCIIpan.}
    See table \ref{tab:tmh_binders_mhc2} for the TMH and epitope counts.
    Note that for smaller proteomes a percentage of zero is likelier.
    \richel{use of top 2\% binders}
  }
  \label{fig:2}
\end{figure}

@@

@@

% Percentage of spots and spots that overlap with a TMH
\input{../bbbq_1_smart/table_f_tmh_2.latex}

@@

Per target proteome (i.e. human, SARS-CoV-2, Mycobacterium tuberculosis),
we collected all 9-mers (for MHC-I) and 15-mers (for MHC-II),
after removing the selenoproteins and proteins that are shorter
than the epitope length.
From these epitopes, per MHC haplotype,
we predicted the IC50 (in nM) using \verb;epitope-prediction; (for MHC-I)
and MHCnuggets (for MHC-II). 
Here, we show the IC50 value per haplotype that
is used to determine if a peptide binds to the haplotype's MHC
for MHC-I (see \ref{tab:ic50_binders_mhc1}) and 
MHC-II (see \ref{tab:ic50_binders_mhc2}).

% tab:ic50_binders_mhc1
\input{../bbbq_1_smart/table_ic50_binders_mhc1_2.latex}

% tab:ic50_binders_mhc2
\input{../bbbq_1_smart/table_ic50_binders_mhc2_2.latex}

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{../bbbq_1_smart/fig_f_tmh_mhc1_2_normalized.png}
  \caption{
    Normalized proportion of MHC-I epitopes overlapping with TMHs
    for human, viral and bacterial proteomes.
    Legend: covid = SARS-CoV-2,
    human = homo sapiens, myco = Mycobacterium tuberculosis
  }
  \label{fig:f_tmh_mhc1_normalized}
\end{figure}

% Label: tab:tmh_binders_mhc1
\input{../bbbq_1_smart/table_tmh_binders_mhc1_2.latex}

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{../bbbq_1_smart/fig_f_tmh_mhc2_2_normalized.png}
  \caption{
    Normalized proportion of MHC-II epitopes overlapping with TMHs
    for human, viral and bacterial proteomes.
    Legend: covid = SARS-CoV-2,
    human = homo sapiens, myco = Mycobacterium tuberculosis
  }
  \label{fig:f_tmh_mhc2_normalized}
\end{figure}

% Label: tab:tmh_binders_mhc2
\input{../bbbq_1_smart/table_tmh_binders_mhc2_2.latex}

>>>>

To determine the percentages of MHC-I and MHC-II epitopes overlap
with TMHs, we used mostly the same analysis as described in \cite{bianchi2017}.
To summarize: from a proteome all possible 9-mers are derived. For each
of these peptides, it is measured if it has been part of a 
TMH (at least overlapping 1 residue), 
as well as if the peptide is a binder to an MHC-I haplotype.
For MHC-II, 14-mers are used, as these are the most occurring
epitope length \cite{bergseng2015different}.

@@

To detect the evolutionary conservation of TMHs, 
we collect human mutations and 
tally their predicted location.
To be more precise, we collected single nucleotide
polymorphisms (SNPs) within the human population
that resulted in the substitution of one amino acid.
As we know the sequence of the protein a SNP acts upon,
we predict that protein's topology and tally if
the SNP occurred in a TMH or not, 
after which we use statistics to determine if there are
more, less or an equal amount of SNPs in TMHs
as a measure of evolutionary conservation.
This workflow is discussed in more detail below and also
depicted in figure \ref{fig:snp_workflow}).

HIERO

\paragraph{Data}
As a data source, multiple
NCBI (\url{https://www.ncbi.nlm.nih.gov/}) databases are used:
'gene' to find the gene names of membrane proteins, 
'dbSNP' (\cite{sherry2001dbsnp}) for SNPs associated with those genes
and 'protein' to obtain the sequence of proteins that SNPs act upon.
The 'dbSNP' contains 650 million 
catalogued non-redundant humane variations (called RefSNPs,
/url{https://www.ncbi.nlm.nih.gov/snp/docs/RefSNP_about/}).
It is estimated that approximately ten percent (ranging from 2 to 36 percent) 
are artifacts (\cite{carlson2003additional, cutler2001high, gabriel2002structure, mitchell2004discrepancies, musumeci2010single, reich2003quality}).

@@

@@

\begin{figure}[!htbp]
  \includegraphics[height=0.9\textheight]{../bbbq_1_smart/results_14-mer/fig_f_tmh_mhc1_2_grid.png}
  \caption{
    Percentage of MHC-I epitopes overlapping with TMHs
    for a human, viral and bacterial proteome.
    The horizontal lines indicate the percentage as expected by chance.
    \richel{Refer to a table with the numbers as expected by chance)}
    See table \ref{tab:tmh_binders_mhc1} for the TMH and epitope counts.
    \richel{Epitope prediction by EpitopePrediction.}.
    Note that for smaller proteomes a percentage of zero is likelier.
    \richel{use of top 2\% binders}
  }
  \label{fig:1}
\end{figure}

@@

\begin{figure}[!htbp]
  \includegraphics[height=0.9\textheight]{../bbbq_1_smart/results_14-mer/fig_f_tmh_mhc2_2_grid.png}
  \caption{
    Percentage of MHC-II epitopes overlapping with TMHs
    for a human, viral and bacterial proteome.
    The horizontal lines indicate the percentage as expected by chance.
    \richel{Refer to a table with the numbers as expected by chance)}
    \richel{Epitope prediction by NetMHCIIpan.}
    See table \ref{tab:tmh_binders_mhc2} for the TMH and epitope counts.
    Note that for smaller proteomes a percentage of zero is likelier.
    \richel{use of top 2\% binders}
  }
  \label{fig:2}
\end{figure}

@@

DRIPs.

Translational staling.

Trans-splicing for epitopes.

@@

% Percentage of spots and spots that overlap with a TMH
\input{../bbbq_1_smart/results_14-mer/table_f_tmh_2.latex}

@@

Per target proteome (i.e. human, SARS-CoV-2, Mycobacterium tuberculosis),
we collected all 9-mers (for MHC-I) and 14-mers (for MHC-II),
after removing the selenoproteins and proteins that are shorter
than the epitope length.
From these epitopes, per MHC haplotype,
we predicted the IC50 (in nM) using \verb;epitope-prediction; (for MHC-I)
and MHCnuggets (for MHC-II). 
Here, we show the IC50 value per haplotype that
is used to determine if a peptide binds to the haplotype's MHC
for MHC-I (see \ref{tab:ic50_binders_mhc1}) and 
MHC-II (see \ref{tab:ic50_binders_mhc2}).

% tab:ic50_binders_mhc1
\input{../bbbq_1_smart/results_14-mer/table_ic50_binders_mhc1_2.latex}

% tab:ic50_binders_mhc2
\input{../bbbq_1_smart/results_14-mer/table_ic50_binders_mhc2_2.latex}

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{../bbbq_1_smart/results_14-mer/fig_f_tmh_mhc1_2_normalized.png}
  \caption{
    Normalized proportion of MHC-I epitopes overlapping with TMHs
    for human, viral and bacterial proteomes.
    Legend: covid = SARS-CoV-2,
    human = homo sapiens, myco = Mycobacterium tuberculosis
  }
  \label{fig:f_tmh_mhc1_normalized}
\end{figure}

% Label: tab:tmh_binders_mhc1
\input{../bbbq_1_smart/results_14-mer/table_tmh_binders_mhc1_2.latex}

@@

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{../bbbq_1_smart/results_14-mer/fig_f_tmh_mhc2_2_normalized.png}
  \caption{
    Normalized proportion of MHC-II epitopes overlapping with TMHs
    for human, viral and bacterial proteomes.
    Legend: covid = SARS-CoV-2,
    human = homo sapiens, myco = Mycobacterium tuberculosis
  }
  \label{fig:f_tmh_mhc2_normalized}
\end{figure}

% Label: tab:tmh_binders_mhc2
\input{../bbbq_1_smart/results_14-mer/table_tmh_binders_mhc2_2.latex}
